<!-- question-type: inclass -->
### Exercise 4: What's Missing? Understanding Joins by Seeing What They Drop

In Exercise 2, we used `left_join()` to combine data from different tables — like firm details, subindustry names, and P/E ratios. That helped us build one data frame with all the information in one place. Now, we'll go a step further and explore what different kinds of joins keep and drop. This can help us spot missing or underrepresented data — and understand our dataset more deeply.

**(a)**.
Let's start by counting how many firms in the ASX200 belong to each subindustry. Use the starter code below to return the three most common industries: 

```{r}
#| eval: false

firm_info |> 
    YOUR_CODE(subind) |> 
    YOUR_CODE(desc(n)) |> 
    YOUR_CODE(3)
```

And, now the three least common:

```{r}
#| eval: false

firm_info |> 
    YOUR_CODE(subind) |> 
    YOUR_CODE((n)) |> 
    YOUR_CODE(3)
```

**(b)**. Are there any subindustries that don’t appear in the ASX200 sample-and so are even less common those that we identify in (a)?

We can use `anti_join()` to find subindustries listed in `subindustry_names` but missing from our firms data. How many subindustries are not included in the ASX200?

```{r}
#| eval: false

YOUR_CODE |> 
    YOUR_CODE(firm_info, by = "subind") |> 
    nrow()
```

**(c)**. What kinds of industries are missing? Why might they be underrepresented in the ASX200?

**(d)**. An `inner_join()` keeps only the firms that have a matching subindustry in the lookup table. Recreate the join you did in Exercise 2(a), but this time use `inner_join()` instead of `left_join()`.

```{r}
#| eval: false

firm_info_inner <- 
    firm_codes |>
    YOUR_CODE(subindustry_names, by = "gsubind")
```

**(e)** Are there any differences in the dataframe in (d) to the one you constructed in 2 (a)? Explain your answer.

**(f)**. Over this exercise and Exercise 2, you've used three different types of joins:

* `left_join()`
* `anti_join()`
* `inner_join()`

Each one gives you a slightly different view of how your data frames relate to each other.

In your own words, answer the following:

1. What does each join keep and what does it drop?
2. Which join is most useful when you want to preserve all rows from one table, even if they don’t match?
3. Which join is best for finding mismatches between two tables?
4. Which join only keeps perfect matches between the two tables?


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- XXXXX

✅   **Core Concepts to Highlight**

List them here


💬 **Suggested In-Class Prompts** (if needed)

List as needed

📌 **Common Misunderstandings**

List as needed

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

```{r}
firm_info |> 
    count(subind) |> 
    arrange(desc(n)) |> 
    head(3)
```

```{r}
firm_info |> 
    count(subind) |> 
    arrange(n) |> 
    head(3)
```

**(b).** 

```{r}
subindustry_names |> 
    anti_join(firm_info, by = "subind") |> 
    nrow()
```

This tells us how many subindustries are not represented in the ASX200 sample — i.e., they appear in the lookup table but not among the top 200 firms.

**(c).**

Look through this output and see if there's any common themes...

```{r}
subindustry_names |> 
    anti_join(firm_info, by = "subind") 
```

**(d)**.

```{r}
firm_info_inner <- 
    firm_codes |>
    inner_join(subindustry_names, by = "gsubind")

firm_info_inner
```

**(e)**. 
The difference is subtle, but if all gsubind codes in `firm_codes` matched a subindustry in the lookup table, then the result will look identical to the `left_join()` output from Exercise 2.

However, if there were any firms with a gsubind code that did not match a code in subindustry_names, those rows would be dropped in the `inner_join()` version but retained (with NA in subind) in the `left_join()`.

Let's see what these are (and re-inforce what `anti_join()` does):

```{r}
firm_info %>%
    anti_join(firm_info_inner, by = "gsubind")
```

This is equivalent to:

```{r}
firm_info |> 
    anti_join(subindustry_names, by = "subind") 
```

We can confirm that this is a case of a gsubind that we don't get a match on in our lookup table:

```{r}
subindustry_names |>
    filter(gsubind == "45101010")
```

**(f).**

1. What does each join keep and drop?

`left_join(x, y)`

* Keeps all rows from x, adds columns from y where there's a match; unmatched rows get NAs.

`inner_join(x, y)`

* Keeps only rows in x that have a match in y (i.e., the intersection).

`anti_join(x, y)`

* Keeps only rows in x that don’t match any row in y.


2. Which join helps preserve all rows from one table?

`left_join()`


3. Which join is best for finding mismatches?

`anti_join()`


4. Which join only keeps complete matches?

`inner_join()`

:::

:::

:::
<!-- END PROFILE:r-solutions -->