<!-- question-type: prepare -->
### Exercise 2: Combining Datasets

In this exercise, we'll combine two or more data frames using joins. Sometimes, joining adds columns. Other times, it adds rows — this is the case when either or all of the data frames include multiple observations for the same join key.

Let's walk through both situations.

To start, let's load the necessary data for this exercise - for two of these data sets, the code below also does some 'light' wrangling (similar to what we did to our data sets in last week's tutorial):

```{r}
#| eval: false

firm_codes <- 
    read_csv("data/asx_200_2024.csv") |> 
    select(gvkey, conm, gsubind)

subindustry_names <-
    read_csv("data/GICS_subindustry.csv")

pe_data <- 
    read_csv("data/pe.csv") |> 
    select(gvkey, fyear, pe) |>
    filter(!is.na(pe)) |> 
    arrange(gvkey, fyear)
```

**(a)**. 
We start with a data frame that lists each firm's subindustry code (`gsubind`), using the largest 200 firms on the ASX in 2024:

```{r}
firm_codes |> 
    head(10)
```

We also have a lookup table that tells us the name of each subindustry:

```{r}
subindustry_names |> 
    head(10)
```

Use a left join to add the subindustry name to each firm in the `firm_codes` dataset using the variable `gsubind` as the join key. The code below gets you started:

```{r}
#| eval: false
firm_info <- 
    YOUR_DATASET_NAME |>
    left_join(YOUR_DATASET_NAME, by = "YOUR_VARIABLE_NAME")

firm_info |> 
    head(10)
```


**(b).** What kind of join did we just perform? To help answer this question, consider the following:

* What kind of thing are we adding: columns or rows?
* Did each `gsubind` only appear once in the dataset `subindustry_names`?
* Does this kind of join have a name? If so, what is it?

**(c)**.

Now let's bring in PE ratio data, which has multiple years per firm:

```{r}
pe_data |> 
    head(10)
```

Use a left join to add the P/E data to each firm in the `firm_info` dataset using the variable `gvkey` as the join key. Use the starter code below to implement this join:

```{r}
#| eval: false
firm_pe_data <- 
    YOUR_DATASET_NAME |>
    YOUR_JOIN_TYPE(YOUR_DATASET_NAME, by = "YOUR_VARIABLE_NAME")

firm_pe_data |>
  head(10) |>
  select(pe, everything())
```

**(d).** What kind of join did we just perform? To help answer this question, consider the following:

* What kind of thing are we adding: columns or rows?
* How many rows are in the dataset `firm_pe_data`? How about `firm_info`?
* Does each `gvkey` in the original table (`firm_info`) now appear once or multiple times in the 'joined' data frame (`firm_pe_data`)? Why?
* Does this kind of join have a name? If so, what is it?

<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Understand how joins work when one table has unique keys and another has repeated keys.

- Practise distinguishing between joins that add columns (many-to-one) and joins that expand rows (one-to-many).

- Gain confidence in using `left_join()` to combine lookup tables and time-series data.

✅   **Core Concepts to Highlight**

Mutating joins: left joins add information while preserving rows from the first table.

Key uniqueness: joins behave differently depending on whether the join key is unique in one or both tables.

Join types in practice: many-to-one (lookup/labels) vs. one-to-many (expanding time dimension).

💬 **Suggested In-Class Prompts** (if needed)

NA

📌 **Common Misunderstandings**

Assuming all joins just “add columns” without realizing they can multiply rows.

Forgetting to check for duplicate keys, leading to unintended row explosion.

Confusing the terminology: e.g., calling a many-to-one join “one-to-one” just because one table looks clean.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). **

```{r}
firm_info <- 
    firm_codes |>
    left_join(subindustry_names, by = "gsubind")

firm_info |>
  head(10)
```

**(b).** This is a mutating join.

What kind of thing are we adding?

* Columns — we're adding a column that contains the subindustry name.

Did each gsubind only appear once in subindustry_names?

* Yes — this is what makes it safe to use as a lookup table.

```{r}
subindustry_names |>
  count(gsubind) |>
  arrange(desc(n))
```

Does this kind of (mutating) join have a name?

* Yes — it’s often called a many-to-one join, because many firms share the same subindustry code, but each code maps to one subindustry name.

```{r}
firm_codes |>
  count(gsubind) |>
  arrange(desc(n))
```

**(c).**

```{r}
firm_pe_data <- 
    firm_info |>
    left_join(pe_data, by = "gvkey")

firm_pe_data |>
  head(10) |>
  select(pe, everything())
```

**(d). ** Again, this is a mutating join.

What kind of thing are we adding?

* Rows (mainly) — each firm now appears once for each year of (pe) data available.
* Note also the PE column is added

How many rows are in `firm_pe_data`? How about `firm_info`?

```{r}
firm_pe_data |> 
    nrow()
```

```{r}
firm_info |> 
    nrow()
```

```{r}
pe_data |>
  nrow()
```


Does each gvkey in `firm_info` now appear once or multiple times? Why?

* Multiple times — once per year of PE data available for that firm.

```{r}
firm_info |>
  count(gvkey) |>
  arrange(gvkey)
```

```{r}
firm_pe_data |>
  count(gvkey) |>
  arrange(gvkey)
```

For example, let's examine `gvkey == "013312"` in `pe_data`:

```{r}
pe_data |>
  filter(gvkey == "013312") 
```

Does this kind of join have a name?
Yes — it's a one-to-many join, because one firm matches to many years in the `pe_data` table.

:::

:::

:::
<!-- END PROFILE:r-solutions -->